---
title: "R First File"
output: 
  pdf_document: default
  html_notebook: default
---

## Welcome to your first program
It might be the case you worked in R before, but it is still the first R file for this class. 
This is going to be our first effort at understanding the R notebook environment and our basic charts. 

## Loading the data
Our data will either be an Excel file or a comma-separated values (.csv) file. 
These are very common and readable by many different programs. 
For most of the R work in this class we will make use of ".csv" files because they make it easier to call the data directly from online. 
There are packages for loading data into R that provide commands. 
Much of what we will use this semester is located in the \em{tidyverse}. 
This is a collection of the most popular collections of code and commands in R. 
It includes collections for reading data, data analysis, and data visualization.

We block off code with the 3 single quotes and then use braces around **r** so it knows what we are using that program at the start and after the commands close it with 3 single quotes.

```{r Package set up and data location}
library(tidyverse)
urlfile="https://raw.githubusercontent.com/flynnecon/datavizf25/refs/heads/main/data/nfldata1.csv"
```
Now that the proper packages are loaded and we point to the site with the data we can focus on getting the data into R.

# Getting the data to usable
To start we are going to read the data in making use of the url.

```{r  Reading data}
nfldata <- read_csv(url(urlfile))
```

## Data Visualizations

We will start with the same chart we created in R, a \textbf{scatterplot}. For this example I make use of the \em(pipe} command, $|>$. This tells R puts whatever is one the left of the \em{pipe} into the first place in the command to the right of the \em{pipe}. The command for generating the chart is $ggplot$.

```{r points scatterplot}
data |> ggplot(aes(x=PF,y=PA)) + geom_point(color="darkgreen")
```

